Ricci-Ollivier Curvature of two Random Walks on the Rooted Phylogenetic Subtree-Prune-Regraft Graph

### 2015-03-28 curvature.tex
curvature distribution observations and a pass through

\usepackage{amsmath,amsthm}
\usepackage{csvsimple}
\usepackage{ifthen}
\usepackage{todonotes}
%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
%\usepackage[right]{showlabels}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{url}
\usepackage{subfigure}
\usepackage[nomarkers,nofiglist]{endfloat}
\let\MYoriglatexcaption\caption
\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
% shift for todonotes
\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
% regular margins
%\usepackage[margin=1in]{geometry}

@@

Ollivier-Ricci curvature provides an intuitive formalization for quantifying the shape of a graph with respect to a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance'') between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.

@@

@@

We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}.
We sampled \ldots \todo{sampling and runtime information}.

\cite{GAP4}

@@

\section{Rooted SPR Neighborhoods}
Having made the connection between curvature values and access times on the SPR graphs, we now consider them theoretically.
In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.

@@

\subsection{Computational investigation and validation}
We computed all of the curvatures for pairs of trees with up to seven leaves to confirm our theoretical work with direct calculation as described before.
\todo{Update once 7 taxon runs have finished}

Indeed we see that for rSPR ... (Figure~\ref{fig:rspr-scatter}). \todo{Chris: can you fill this in?}

\section{Conclusion and future work}
In summary, we have extended knowledge about the the phylogenetic rSPR graph underlying phylogenetic inference beyond previous work concerning graph diameter and vertex degree.
We done so by developing theoretical and computational frameworks to compute Ricci-Ollivier curvature of the rSPR graph.
We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.

>>>>

\usepackage{amsmath,amsthm}
\usepackage{csvsimple}
\usepackage{ifthen}
\usepackage{todonotes}
%\usepackage[notref,notcite]{showkeys} % show keys for eqs, etc.
%\usepackage[right]{showlabels}
\usepackage{cite}
\usepackage{enumitem}
\usepackage{url}
\usepackage{subfigure}
\usepackage[nomarkers,nofiglist,notablist]{endfloat}
\let\MYoriglatexcaption\caption
\renewcommand{\caption}[2][\relax]{\MYoriglatexcaption[#2]{#2}}
% shift for todonotes
\usepackage[left=0.5in,right=1.5in,top=1in,bottom=1in,marginparwidth=1.25in]{geometry}
% regular margins
%\usepackage[margin=1in]{geometry}

@@

Ollivier-Ricci curvature provides an intuitive formalization for quantifying the shape of a graph with respect to a random walk (or indeed any metric space, see~\cite{Ollivier2009-bw}).
Let $m_x$ and $m_y$ be probability densities of a specified random walk starting at points $x$ and $y$ of a graph $G = (V,E)$, respectively.
The Wasserstein distance~\cite{Villani2003-wv} (equivalently $L^1$ transportation, or ``earth movers distance''~\cite{rubner2000earth}) between $m_x$ and $m_y$ is the minimum amount of ``work'' required to move $m_x$ to $m_y$ following the graph, that is
 $$ W_1(m_x, m_y) := \min_{\xi \in \Pi(m_x, m_y)} \sum \dspr{x,y} \xi(x,y),$$
where $\Pi(m_x, m_y)$ is the set of of assignments of probability from $m_x$ to pairs of vertices in $V \times V$.

@@

We saw a wide variation in curvature values among tanglegrams (Figure~\ref{fig:rspr-scatter}).~\todo{EM: add 7-taxa graph when ready}
Curvature values tended to increase with increasing SPR distance.
In addition, their variance decreased with increasing distance.
Minimum curvature values for a given treespace were observed between neighboring trees, while maximum curvature values were observed between trees at maximum distance or one rSPR move closer than the maximum.
This suggests that the increased difficulty of moving between trees in a random walk due to distance may be tempered somewhat by curvature in the highly connected rSPR treespaces.

Larger treespaces tended to have smaller curvature values.
Indeed, the 7-leaf treespace contained adjacent pairs of trees with negative curvature.
Such pairs are possible bottlenecks in phylogenetic searches which may be exarcebated by likelihood or branch length constraints.

@@

We implemented this procedure in the C++ package \texttt{random\_spr\_walk}~\cite{random_spr_walk}. 
We sampled a 200,000-iteration random walk on the 4-leaf rSPR graph and a 50,000-iteration random walk on the 5-leaf rSPR graph.
This required XXX and YYY time respectively.~\todo{EM: can you fill in the time this required?}

@@

\section{Rooted SPR Neighborhoods}
Having made the connection between curvature values and access times on the SPR graphs, we now consider curvature theoretically.
In the absence of a bias provided by a likelihood function, curvature for the random walks we consider is entirely determined by the degree of vertices in the graph.
Thus, in order to provide bounds on curvature, we will begin by bounding differences between degrees.

@@

\section{Conclusion and future work}
In summary, we have extended knowledge about the the phylogenetic rSPR graph underlying phylogenetic inference beyond previous work concerning graph diameter and vertex degree.
We did so by developing theoretical and computational frameworks to bound and compute Ricci-Ollivier curvature of the rSPR graph.
We find that curvature, along with degree and distance, determine the early dynamics of hitting times for random walks.
Moreover, we proved that rSPR graph degree changes depend quadratically on the product of the size of the regrafted subtree and its change in depth and that rSPR treespace tends toward flatness with respect to SPR moves that move small subtrees.
